---
title: "Activity 2: Reshaping data"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

# GDP data

Run the following code to import data on GDP for each country (you may need to install the `googlesheets4` package):

```{r, message=F, warning=F}
library(tidyverse)
library(googlesheets4)
gs4_deauth()
gdp_data <- read_sheet("https://docs.google.com/spreadsheets/d/1RctTQmKB0hzbm1E8rGcufYdMshRdhmYdeL29nXqmvsc/pub?gid=0")
```

Looking at the data, we can see that there is a column for the GDP in each year, and the first column contains country information:

```{r}
head(gdp_data)
```

:::{.question}
#### Question 1
Use the `pivot_longer` function to reshape the GDP data so it looks like this:

```{r, echo=F}
gdp_data |>
  rename(country = starts_with("Income")) |>
  pivot_longer(cols = -country,
               names_to = "year",
               values_to = "gdp",
               values_drop_na = T)
```

* There should be no NAs
* The first column should be renamed `country`
* There should be separate columns for `year` and `gdp`
:::

# WHO data

The `tidyr` package contains a dataset called `who2` which contains information on tuberculosis counts for each country in each year, broken down by method of diagnosis, sex, and age group:

```{r}
who2
```

Columns like `sp_m_1524` contain the number of tuberculosis cases. The column names contain information about the method of diagnosis (`rel`, `sn`, `sp`, or `ep`), sex (`f` or `m`), and age group (`014` = 0-14 years, `1524` = 15-24 years, etc.).

Ideally, we would like these variables to each have their own column. 

:::{.question}
#### Question 2
Reshape the `who2` data so it looks like this:

```{r, echo=F}
who2 |>
  pivot_longer(cols = -c(country, year),
               names_to = c("diagnosis", "sex", "age_group"),
               names_sep = "_",
               values_to = "count",
               values_drop_na = T)
```

* There should be no NAs
* There should be separate columns for `diagnosis`, `sex`, and `age_group`
* The `count` column contains the number of cases
:::

# household data

The `household` data from the `tidyr` package contains information about five different households and the children in each household:

```{r}
household
```

:::{.question}
#### Question 3
Using `".value"` in the `names_to` argument of the `pivot_longer` function, reshape the `household` data to look like this:

```{r, echo=F}
household |>
  pivot_longer(cols = -family,
               names_to = c(".value", "child"),
               names_sep = "_",
               values_drop_na = T)
```
:::


*Questions adapted from Jo Hardin, Math 154 Computational Statistics*


