---
title: "Activity 5: More data wrangling practice questions"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

```{r, include=F, message=F, warning=F}
library(tidyverse)
```


## Reading data wrangling code

Here are two small datasets, `df1` and `df2`:

```{r, include=F}
set.seed(3)

df1 <- data.frame(id = 1:5, 
                  x = sample(1:10, 5, replace=T),
                  y = sample(1:10, 5, replace=T),
                  z = sample(1:10, 5, replace=T))

df2 <- data.frame(id = 3:6,
                  a = sample(1:10, 4, replace=T),
                  b = sample(1:10, 4, replace=T))
```

```{r}
df1

df2
```


For each of the following chunks of code, write down the output or explain why it will cause an error.


1.

```{r, eval=F}
df1 |>
  left_join(df2, join_by(id))
```

2.

```{r, eval=F}
df1 |>
  inner_join(df2, join_by(id))
```

3.

```{r, eval=F}
df1 |>
  group_by(z) |>
  summarize(max_b = max(b))
```


4.

```{r, eval=F}
df1 |>
  select(x, y) |>
  pivot_longer(cols = -id,
               names_to = "measurement",
               values_to = "value")
```

5.

```{r, eval=F}
df1 |>
  select(id, x, y) |>
  pivot_longer(cols = -id,
               names_to = "measurement",
               values_to = "value") |>
  filter(id %in% c(1, 2, 3))
```

6.

```{r, eval=F}
df1 |>
  left_join(df2, join_by(id)) |>
  mutate(new_var = x + a) |>
  group_by(z) |>
  summarize(mean_new_var = mean(new_var))
```


7.

```{r, eval=F}
df1 |>
  left_join(df2, join_by(id)) |>
  mutate(new_var = x + a) |>
  group_by(z) |>
  summarize(mean_new_var = mean(new_var, na.rm=T)) |>
  summarize(mean_b = mean(b))
```

## Writing data wrangling code

In each of the questions below, write code to produce the output from the original data.

8. Original data:

```{r, echo=F}
ex_df <- data.frame(
  id = c(1, 2, 3),
  x_1 = c(3, 1, 4),
  x_2 = c(5, 8, 9),
  y_1 = c(0, 1, 2),
  y_2 = c(2, 7, 9)
)

ex_df
```

Output:

```{r, echo=F}
ex_df |>
  pivot_longer(cols = -id, names_to = c("group", "obs"), names_sep = "_")
```

9. Original data:

```{r, echo=F}
ex_df <- data.frame(
  id = c(1, 1, 2, 2, 3, 3),
  group = c("x", "y", "x", "y", "x", "y"),
  value = sample(1:6, replace=T)
)

ex_df
```

Output:

```{r, echo=F}
ex_df |>
  pivot_wider(id_cols = id, names_from = group, values_from = value)
```


10. Original data:

```{r, echo=F}
df1 <- data.frame(
  id = c(1, 2, 3),
  x = c(7, 9, 13)
)

df2 <- data.frame(
  id = c(1, 2, 4),
  y = c(10, 12, 14)
)
```

```{r}
df1

df2
```


Output:

```{r, echo=F}
df1 |>
  left_join(df2, join_by(id))
```


11. Original data:

```{r}
df1

df2
```


Output:

```{r, echo=F}
df1 |>
  inner_join(df2, join_by(id))
```

12. Original data:

```{r, echo=F}
df1 = data.frame(
  a_x = 1,
  a_y = 2,
  b_x = 2,
  b_y = 3
)

df2 = data.frame(
  id = c("a", "b"),
  z = c(4, 5)
)
```

```{r}
df1

df2
```


```{r, echo=F}
df1 |> 
  pivot_longer(cols = -c(), names_to = c("id", ".value"), names_sep = "_") |>
  left_join(df2, join_by(id))
```