---
title: "Activity 3: Function and iteration practice questions"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

## Practice with iteration in R

For each of the following questions, either write the output of the code, or explain why it gives an error. (Some questions will run correctly, others will cause errors)

1.

```{r, eval=F}
output <- rep(NA, 10)
for(i in 1:5){
  output[i] <- i
}

output[6]
```

2.

```{r, eval=F}
output <- rep(0, 10)
for(i in 1:10){
  output[i] <- i
}

output[6]
```

3. 

```{r, eval=F}
output <- rep(0, 10)
for(i in 1:10){
  output[i] <- i
}

output[11]
```

4. 

```{r, eval=F}
output <- rep(1, 10)
for(i in 2:10){
  output[i] <- i + output[i-1]
}

output[5]
```

5. 

```{r, eval=F}
output <- rep(1, 10)
for(i in 2:10){
  output[i] <- i + output[i+1]
}

output[5]
```

6.

```{r, eval=F}
x <- list()
for(i in 1:10){
  x[i] <- i
}
x[2] + 1
```

7.

```{r, eval=F}
mat <- matrix(0, nrow=5, ncol=3)
for(i in 1:5){
  for(j in 1:3){
    mat <- i + j
  }
}

mat
```

8.

```{r, eval=F}
mat <- matrix(0, nrow=5, ncol=3)
for(i in 1:5){
  for(j in 1:3){
    mat[j, i] <- i + j
  }
}

mat
```

9.

```{r, eval=F}
mat <- matrix(1, 3, 3)
for(i in 2:3){
  for(j in 2:3){
    mat[i,j] <- mat[i-1, j-1] + mat[i, j-1]
  }
}

mat
```

10.

```{r, eval=F}
f1 <- function(n, groups){
  x <- matrix(1, nrow=n, ncol=n)
  unique_groups = unique(groups)
  means <- matrix(nrow = length(unique_groups), ncol = n)
  for(i in 1:length(unique_groups)){
    means[i,] <- colMeans(x[groups == unique_groups[i],])
  }
  
  return(means)
}

f1(5, groups = c(1, 1, 2, 2, 2))
```

## Practice with functions

For each question, predict what will happen when the code is run. Then run the code and check whether your prediction was correct.

11. 

```{r, eval=F}
x <- 10

g02 <- function(x){
  x <- x + 1
  return(x)
}

g02(x)
x + 1
```

12.

```{r, eval=F}
x <- 10

g02 <- function(x){
  x <- x + 1
  return(x)
}

x <- g02(x)
x + 1
```

13. 

```{r, eval=F}
g02 <- function(y){
  y <- y - 1
  return(y)
}

g02(g02(20))
```

14. 

```{r, eval=F}
x <- 10
test_fun <- function(x = 11){
  return(x)
}
test_fun()
x
```

15. 

```{r, eval=F}
x <- 10
test_fun <- function(y = 11){
  return(x + 1)
}
test_fun()
x
```

16. 

```{r, eval=F}
x <- 10
test_fun <- function(y = 11){
  x <- x + 1
  return(x + 1)
}
test_fun()
x
```

17. 

```{r, eval=F}
x <- 10
test_fun <- function(x = 11){
  x <- x + 1
  return(x + 1)
}
test_fun()
x
```

18. 

```{r, eval=F}
x <- 10
test_fun <- function(x = 11){
  x <- x + 1
  return(x + 1)
}
x <- test_fun(x)
x
```

19.

```{r, eval=F}
f1 <- function(x = 1){
  return(x + 1)
}
g1 <- function(x){
  return(f1() + x)
}

f1(g1(3))
```

## Practice with anonymous functions

20. Use the `integrate()` and an anonymous function to find the area under the curve for the following functions:

* `y = x^2 - x` for x in $[0, 1]$
* `y = sin(x) + cos(x)` for x in $[-\pi, \pi]$
* `y = exp(x)/x` for x in $[10, 20]$