---
title: "Activity 1: Data wrangling across columns"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

# Data

For this class activity, we will create our own small toy dataset, so that we can explore data wrangling functions with a dataset that is easy to view and work with.

Run the following code in R to create the toy dataset:

```{r, message=F, warning=F}
library(tidyverse)

example_df <- data.frame(
  x1 = c(1, 2, 3),
  x2 = c("a", "b", "c"),
  x3 = c(5, 1, 2),
  y1 = c(0, 9, 2),
  y2 = c(2, 7, 9),
  z = c(0, 0, 0)
)

example_df
```

# Efficiently selecting multiple columns

Suppose we want to `select` only the columns `y1` and `y2` in this toy dataset. A simple way to code this is:

```{r}
example_df |>
  select(y1, y2)
```

However, with multiple columns, it can be difficult or tedious to list them all out by hand! Instead, we can `select` columns that meet certain *criteria*.

For example, we can use the `starts_with` function to select all columns which begin with `y`:

```{r}
example_df |>
  select(starts_with("y"))
```


:::{.question}
#### Question 1
Modify the code above to select the columns `x1`, `x2`, and `x3`, *without* listing them explicitly.
:::

What if we want only the columns which contain characters? (e.g. `"a"`, `"b"`, etc.) The `where` function returns the columns *where* a given condition is true:

```{r}
example_df |>
  select(where(is.character))
```

:::{.question}
#### Question 2
Modify the code above to select only the *numeric* columns (hint: use `is.numeric`).
:::

:::{.question}
#### Question 3
Now select only the numeric columns which start with `x` (hint: use `&`)
:::

# Applying functions to multiple columns

Suppose I want the mean of both `y1` and `y2`. One option, of course, is to list them explicitly in `summarize`:

```{r}
example_df |>
  summarize(y1_mean = mean(y1),
            y2_mean = mean(y2))
```

However, as before, this gets tedious with multiple columns! Instead, we can use `across` to apply a function *across* multiple columns:

```{r}
example_df |>
  summarize(across(starts_with("y"), mean))
```

What if I want to apply multiple functions? I specify all the functions I want in a *list*:

```{r}
example_df |>
  summarize(across(starts_with("y"), list(mean, sd)))
```

However, the names here aren't very useful! We can fix this by naming the elements of our list:

```{r}
example_df |>
  summarize(across(starts_with("y"), list("mean" = mean, "sd" = sd)))
```

:::{.question}
#### Question 4
Modify the code above to calculate the median and IQR for the numeric columns which start with `x`. Do not list the columns explicitly.
:::
