---
title: "Activity 3 solutions"
format: 
  html:
    embed-resources: true
editor: source
author: "Ciaran Evans"
---

```{r include=F}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(warning = F)
knitr::opts_chunk$set(error = T)
```

1. No -- we have multiple years for each country in the data. I would expect observations for the same country in different years to be strongly correlated.

2.

```{r}
library(tidyverse)
library(gapminder)

gapminder_2007 <- gapminder |>
  filter(year == 2007)
```

3.

```{r}
m1 <- lm(lifeExp ~ gdpPercap, data = gapminder_2007)
```


```{r}
gapminder_2007 |>
  mutate(yhat = predict(m1),
         residuals = residuals(m1)) |>
  ggplot(aes(x = yhat, y = residuals)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, 
              color = "blue", lwd = 1.2) +
  labs(x = "Predicted life expectancy",
       y = "Residual") +
  theme_bw()
```

```{r}
gapminder_2007 |>
  mutate(residuals = residuals(m1)) |>
  ggplot(aes(sample = residuals)) +
  geom_qq() +
  geom_qq_line() +
  labs(x = "Theoretical normal quantiles",
       y = "Observed residual quantiles") +
  theme_classic()
```

The assumption of a linear relationship between GDP per capita and life expectancy definitely does *not* seem reasonable. We will need to address that violation, then we can re-assess the other assumptions.

4.

```{r}
m2 <- lm(lifeExp ~ log(gdpPercap), data = gapminder_2007)
```


5.

```{r}
gapminder_2007 |>
  mutate(yhat = predict(m2),
         residuals = residuals(m2)) |>
  ggplot(aes(x = yhat, y = residuals)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, 
              color = "blue", lwd = 1.2) +
  labs(x = "Predicted life expectancy",
       y = "Residual") +
  theme_bw()
```

```{r}
gapminder_2007 |>
  mutate(residuals = residuals(m2)) |>
  ggplot(aes(sample = residuals)) +
  geom_qq() +
  geom_qq_line() +
  labs(x = "Theoretical normal quantiles",
       y = "Observed residual quantiles") +
  theme_classic()
```

While the diagnostics are not perfect, the shape assumption looks much better than it did before.

6. 

```{r}
coef(m2)

coef(m2)[1] + coef(m2)[2]*log(20000)
```
The estimated life expectancy is 76.28 years.







